---
title: "01-get-data"
author: "Sarah Tanja"
date: "4/11/2023"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)
```

# Getting RNAseq FASTQ Data Files

prereqs : have sequence data files deposited on NCBI

## 1. Get RNAseq files from National Center for Biotechnology Information (NCBI) Sequence Read Archive (SRA)

👀Checkout some background info on [The Sequence Read Archive (SRA)](https://linsalrob.github.io/ComputationalGenomicsManual/Databases/SRA.html#:~:text=A%20Study%20(SRP)%20has%20one,want%20to%20download%20from%20NCBI.) and the [SRA factsheet](https://www.ncbi.nlm.nih.gov/core/assets/sra/files/Factsheet_SRA.pdf)

👀Checkout the [Bioconductor SRAdb package](https://bioconductor.org/packages/release/bioc/html/SRAdb.html)

```{r install packages}
if ("BiocManager" %in% rownames(installed.packages()) == 'FALSE') install.packages("BiocManager") 

if ("SRAdb" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("SRAdb")
```

```{r load libraries}
library(SRAdb)
library(tidyverse)
```

First we want to get an idea of the files we are downloading and the samples that generated the data. We will start by looking at the metadata for the samples.

##### Get sample metadata :

```{bash}
# navigate to data directory
cd ../data

# download metadata from the git repo in raw
curl -O https://raw.githubusercontent.com/AHuffmyer/EarlyLifeHistory_Energetics/master/Mcap2020/Data/TagSeq/Sample_Info.csv

```

```{r}
#pull data into R and rename it metadata 
metadata <- read_csv("../data/Sample_Info.csv")
```

```{r}
#look at the metadata
head(metadata)
```

There are 39 samples in this dataset (AH1 - AH39).

##### All the files should look like:

+-------------------+-------------------------------+-----------------------------------+
| **TagSeq seq ID** | **TagSeq File 1**             | **TagSeq biosample accession \#** |
+===================+===============================+===================================+
| AH1               | AH1_S32_L002_R1_001.fastq.gz  | SRR22293483                       |
+-------------------+-------------------------------+-----------------------------------+
| AH2               | AH2_S33_L002_R1_001.fastq.gz  | SRR22293482                       |
+-------------------+-------------------------------+-----------------------------------+
| AH3               | AH3_S34_L002_R1_001.fastq.gz  | SRR22293471                       |
+-------------------+-------------------------------+-----------------------------------+
| AH4               | AH4_S35_L002_R1_001.fastq.gz  | SRR22293460                       |
+-------------------+-------------------------------+-----------------------------------+
| AH5               | AH5_S36_L002_R1_001.fastq.gz  | SRR22293450                       |
+-------------------+-------------------------------+-----------------------------------+
| AH6               | AH6_S37_L002_R1_001.fastq.gz  | SRR22293449                       |
+-------------------+-------------------------------+-----------------------------------+
| AH7               | AH7_S38_L002_R1_001.fastq.gz  | SRR22293448                       |
+-------------------+-------------------------------+-----------------------------------+
| AH8               | AH8_S39_L002_R1_001.fastq.gz  | SRR22293447                       |
+-------------------+-------------------------------+-----------------------------------+
| AH9               | AH9_S40_L002_R1_001.fastq.gz  | SRR22293446                       |
+-------------------+-------------------------------+-----------------------------------+
| AH10              | AH10_S41_L002_R1_001.fastq.gz | SRR22293445                       |
+-------------------+-------------------------------+-----------------------------------+
| AH11              | AH11_S42_L002_R1_001.fastq.gz | SRR22293481                       |
+-------------------+-------------------------------+-----------------------------------+
| AH12              | AH12_S43_L002_R1_001.fastq.gz | SRR22293480                       |
+-------------------+-------------------------------+-----------------------------------+
| AH13              | AH13_S44_L002_R1_001.fastq.gz | SRR22293479                       |
+-------------------+-------------------------------+-----------------------------------+
| AH14              | AH14_S45_L002_R1_001.fastq.gz | SRR22293478                       |
+-------------------+-------------------------------+-----------------------------------+
| AH15              | AH15_S46_L002_R1_001.fastq.gz | SRR22293477                       |
+-------------------+-------------------------------+-----------------------------------+
| AH16              | AH16_S47_L002_R1_001.fastq.gz | SRR22293476                       |
+-------------------+-------------------------------+-----------------------------------+
| AH17              | AH17_S48_L002_R1_001.fastq.gz | SRR22293475                       |
+-------------------+-------------------------------+-----------------------------------+
| AH18              | AH18_S49_L002_R1_001.fastq.gz | SRR22293474                       |
+-------------------+-------------------------------+-----------------------------------+
| AH19              | AH19_S50_L002_R1_001.fastq.gz | SRR22293473                       |
+-------------------+-------------------------------+-----------------------------------+
| AH20              | AH20_S51_L002_R1_001.fastq.gz | SRR22293472                       |
+-------------------+-------------------------------+-----------------------------------+
| AH21              | AH21_S52_L002_R1_001.fastq.gz | SRR22293470                       |
+-------------------+-------------------------------+-----------------------------------+
| AH22              | AH22_S53_L002_R1_001.fastq.gz | SRR22293469                       |
+-------------------+-------------------------------+-----------------------------------+
| AH23              | AH23_S54_L002_R1_001.fastq.gz | SRR22293468                       |
+-------------------+-------------------------------+-----------------------------------+
| AH24              | AH24_S55_L002_R1_001.fastq.gz | SRR22293467                       |
+-------------------+-------------------------------+-----------------------------------+
| AH25              | AH25_S56_L002_R1_001.fastq.gz | SRR22293466                       |
+-------------------+-------------------------------+-----------------------------------+
| AH26              | AH26_S57_L002_R1_001.fastq.gz | SRR22293465                       |
+-------------------+-------------------------------+-----------------------------------+
| AH27              | AH27_S58_L002_R1_001.fastq.gz | SRR22293464                       |
+-------------------+-------------------------------+-----------------------------------+
| AH28              | AH28_S59_L002_R1_001.fastq.gz | SRR22293463                       |
+-------------------+-------------------------------+-----------------------------------+
| AH29              | AH29_S60_L002_R1_001.fastq.gz | SRR22293462                       |
+-------------------+-------------------------------+-----------------------------------+
| AH30              | AH30_S61_L002_R1_001.fastq.gz | SRR22293461                       |
+-------------------+-------------------------------+-----------------------------------+
| AH31              | AH31_S62_L002_R1_001.fastq.gz | SRR22293459                       |
+-------------------+-------------------------------+-----------------------------------+
| AH32              | AH32_S63_L002_R1_001.fastq.gz | SRR22293458                       |
+-------------------+-------------------------------+-----------------------------------+
| AH33              | AH33_S64_L002_R1_001.fastq.gz | SRR22293457                       |
+-------------------+-------------------------------+-----------------------------------+
| AH34              | AH34_S65_L002_R1_001.fastq.gz | SRR22293456                       |
+-------------------+-------------------------------+-----------------------------------+
| AH35              | AH35_S66_L002_R1_001.fastq.gz | SRR22293455                       |
+-------------------+-------------------------------+-----------------------------------+
| AH36              | AH36_S67_L002_R1_001.fastq.gz | SRR22293454                       |
+-------------------+-------------------------------+-----------------------------------+
| AH37              | AH37_S68_L002_R1_001.fastq.gz | SRR22293453                       |
+-------------------+-------------------------------+-----------------------------------+
| AH38              | AH38_S69_L002_R1_001.fastq.gz | SRR22293452                       |
+-------------------+-------------------------------+-----------------------------------+
| AH39              | AH39_S70_L002_R1_001.fastq.gz | SRR22293451                       |
+-------------------+-------------------------------+-----------------------------------+

### Obtain Run (SRR) numbers

> *"Results are called runs (SRR). Runs comprise the data gathered for a sample or sample bundle and refer to a defining experiment."*

First, we need to obtain RNAseq files from NCBI from the project [PRJNA900235](https://www.ncbi.nlm.nih.gov/bioproject/?term=PRJNA900235). In order to do this we need a list of SRR numbers that identify the specific sequence files for RNAseq. Go to the [SRA links](https://www.ncbi.nlm.nih.gov/sra?LinkName=bioproject_sra_all&from_uid=900235) for BioProject 900235 and select all the RNAseq files of interest. In the upper right hand corner, I selected "Send to" and "File". This outputs a .txt file with a list of all SRR numbers.
