---
title: "02-build-reference-genome"
date: "`r format(Sys.time(), '%d %B, %Y')`"  
format:
  html:
    df-print: paged
    toc: true
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    code-fold: show
    code-tools: true
    code-copy: true
    highlight-style: arrow
    code-overflow: wrap
    theme:
      light: sandstone
      dark: vapor
  gfm: default
---

# Download Annotated Reference Genome for *Montipora capitata*

Deep Dive project with genomes of interest: <https://github.com/urol-e5/deep-dive>

*Montipora capitata* Genome version V3, Rutgers University: <http://cyanophora.rutgers.edu/montipora/>

Genome publication: <https://academic.oup.com/gigascience/article/doi/10.1093/gigascience/giac098/6815755>

Nucleotide Coding Sequence (CDS): <http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.cds.fna.gz>

This code grabs the *Montipora capitata* fasta file (rna.fna) of genes.

```{r, engine='bash'}
# change to work in data directory
cd ../data
# download the rna.fna file to data directory from the gannet server
curl -O http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.cds.fna.gz

wget http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.assembly.fasta.gz

wget http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.gff3.gz
```

Obtain `Montipora_capitata_HIv3.genes_fixed.gff3` file by downloading from GitHub.

```{r, engine='bash'}
cd ../data
wget https://github.com/AHuffmyer/EarlyLifeHistory_Energetics/raw/master/Mcap2020/Data/TagSeq/Montipora_capitata_HIv3.genes_fixed.gff3.gz
```

This was generated by running the original gff file `Montipora_capitata_HIv3.genes.gff3.gz` through this script in R: <https://github.com/AHuffmyer/EarlyLifeHistory_Energetics/blob/master/Mcap2020/Scripts/TagSeq/Genome_V3/fix_gff_format.Rmd>

Unzip gff and genome file

```{r, engine='bash'}
cd ../data
gunzip Montipora_capitata_HIv3.genes.gff3.gz
gunzip Montipora_capitata_HIv3.genes_fixed.gff3.gz
gunzip Montipora_capitata_HIv3.assembly.fasta.gz
```

# 
